Fractional Types

### 2013-02-20 frac.tex
intro to sec3

The language {{Pi}} models isomorphisms of values rather well.  However,
purposefully, it has no distinguished notion of \emph{input} or
\emph{output}; more precisely, because of strict preservation of information,
these two concepts coincide.  This is the root of reversibility. But if we
want to model functions of any flavor as values, we need to differentiate
between these notions.  Our idea, inspired by computational
dualities~\cite{Filinski:1989:DCI:648332.755574,Curien:2000},
polarity~\cite{Girard87tcs,10.1109/LICS.2010.23}, and the categorical notion
of \emph{compact
  closure}~\cite{Selinger:2007:DCC:1229185.1229207,Abramsky:2004:CSQ:1018438.1021878},
is to introduce a \emph{formal dual} for our types.  We thus consider a type
in a negative position to possess negative information, as it is really a
request for information.  Since information is logarithmic, this means that a
request for information should behave like a \emph{fractional type}.

@@

Syntactically, the extension with fractional types is simple.

\begin{definition}[Syntax of {{PiTF}}]
\label{def:PiT}
We extend our previous definition with:
%subcode{bnf} include main
% value types, b ::= ... | 1/b
% values, v ::= ... | 1/v
%
% base iso ::= ... | eta*_b | eps*_b
where:
{{eta*_b : 1 <-> 1/b * b : eps*_b}}
\end{definition}

For a given type {{b}}, the values of type {{1/b}} are of the form {{1/v}}
where {{v : b}}. Note that {{1/v}} is purely formal, as is {{1/b}}.  However,
we think of a value like {{1/left ()}} as a \emph{first-class constraint}
that can only be satisfied if it is matched with an actual value {{left ()}}
or as representing the absence of some information (i.e., \emph{negative
  information}), that can only be reconciliated by {{left ()}}. The
combinator {{eta*_b : 1 <-> 1/b * b}} represents a fission point which
creates --- out of no information --- an equal amount of negative and
positive information. Symmetrically, the combinator 
{{eps*_b : 1/b * b <-> 1}} 
matches up an equal amount of negative and positive information producing no
residual information.  The best analogy is probably that of vectors and
co-vectors, also know as column vectors and row vectors (respectively) where
the multiplication of the row vector by the column vector is a scalar. In the
finite dimensional case, they are isomorphic, but they are nevertheless quite
different, being $(1,0)$ and $(0,1)$ tensors (respectively).  In other words,
since we are explicitly interested in \emph{isomorphisms}, we should not
conflate negative-information values with positive-information values just
because they happen to be isomorphic. (See Sec.~\ref{sec:cat} for a more
thorough discussion about the meaning of fractional types and values in both an
algebraic and categorical sense.) For now, we proceed with an operational 
understanding of the language.

The semantics we provide is simple: each combinator {{c : b1 <-> b2}} is
interpreted as a \emph{relation} between the sets denoted by {{b1}} and
{{b2}} respectively.

@@

More generally, the semantics of {{PiTF}} using sets and relations, i.e., in
the category of finite sets and relations suggests a few additional immediate
properties.  The category of finite sets and relations is an example of a
\emph{biproduct dagger compact closed category} which are the categories
introduced by Abramsky and Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878}
and by Selinger~\cite{Selinger:2007:DCC:1229185.1229207} to model quantum
computation.\footnote{The other common example of these categories is that of
  finite dimensional Hilbert spaces.} Two of the most relevant features of
these categories for our purposes are that they support an adjoint for each
morphism (to model reversibility) and that they support higher-order relation
(built from the categorical dagger which is modeled by our fractional
types). The latter is intuitively clear as any relation between $A$ and $B$
can be represented as a subset of {{A * B}}, i.e., as a value. We return to
such properties of {{PiTF}} in Sec.~\ref{sec:cat} after we present
programming examples that illustrate {{PiTF}} and its ability to express
various higher-order notions.

>>>>

The language {{Pi}} models isomorphisms of values rather well.  However,
purposefully, it has no distinguished notion of \emph{input} or
\emph{output}; more precisely, because of strict preservation of information,
these two concepts coincide.  This is the root of reversibility. But if we
want to model functions of any flavor as values, we need to differentiate
between these notions. We also clearly need to relax the strict separation
between values and isomorphisms as the notion of ``higher-order'' requires
that isomorphisms be represented as values. Our idea, inspired by
computational dualities~\cite{Filinski:1989:DCI:648332.755574,Curien:2000},
polarity~\cite{Girard87tcs,10.1109/LICS.2010.23}, and the categorical notion
of \emph{compact
  closure}~\cite{Selinger:2007:DCC:1229185.1229207,Abramsky:2004:CSQ:1018438.1021878},
is to introduce a \emph{formal dual} for our types.  We thus consider a type
in a negative position to possess negative information, as it is really a
request for information.  Since information is logarithmic, this means that a
request for information should behave like a \emph{fractional type}.

@@

The extension of the sets of types and values is simple:

%subcode{bnf} include main
% value types, b ::= ... | 1/b
% values, v ::= ... | 1/v

For a given type {{b}}, the values of type {{1/b}} are of the form {{1/v}}
where {{v : b}}. Note that {{1/v}} is purely formal, as is {{1/b}}. There is
however have some informal intuition about the \emph{meaning} of such
fractional types and values. (See Sec.~\ref{sec:cat} for more discussion
about this point.) From both the programming perspective and from the
categorical perspective we expect the fractional types to satisfy the
following isomorphism witnessed by two new combinators {{eta*_b}} and
{{eps*_b}}:

{{eta*_b : 1 <-> 1/b * b : eps*_b}}

From a programming perspective, we think of a value {{1/v}} as a
\emph{first-class constraint} that can only be satisfied if it is matched
with an actual value {{v}} or as a \emph{pattern} representing the absence of
some information of a specific shapre (i.e., \emph{negative information}),
that can only be reconciliated by an actual value {{v}}. The combinator
{{eta*_b : 1 <-> 1/b * b}} thus represents a fission point which creates ---
out of no information --- an equal amount of negative and positive
information. Symmetrically, the combinator {{eps*_b : 1/b * b <-> 1}} matches
up an equal amount of negative and positive information producing no residual
information. From the categorical perspective, the extended language {{PiTF}}
needs to have a semantics in a \emph{compact closed category} which are the
categories introduced by Abramsky and
Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878} and by
Selinger~\cite{Selinger:2007:DCC:1229185.1229207} to model quantum
computation. Two of the most relevant features of these categories for our
purposes are that they support an adjoint for each morphism (to model
reversibility) and that they support higher-order relation (built from the
categorical dagger which is modeled by our fractional types). The main
requirement on the dagger operation is the isomorphism above witnessed by
{{eta*_b}} and {{eps*_b}}.

One can think of several concrete compact closed categories: the first and
most intuitive is the category of finite sets and relations. In that case,
the dagger operation (i.e., the fractional type) collapses by defining
{{1/b}} to be {{b}}. This is still an interesting category as any relation is
trivially reversible and more importantly any relation between {{b1}} and
{{b2}} can be represented as a subset of {{b1 * b2}}, i.e., as a value. A
more interesting example is the category of finite dimensional vector spaces
and linear maps (or even the category of finite dimensional Hilbert
spaces). In such categories the dagger operation takes a vector space
(consisting of ``column vectors'') to the dual vector space (consisting of
``row vectors''). These vector spaces are isomorphic, but they are
nevertheless quite different, being $(1,0)$ and $(0,1)$ tensors
(respectively).  In other words, the category provides a more refined model
in which isomorphic negative-information values and positive-information are
not identified.

In this section, we provide a simple semantics of {{PiTF}} in the category of
finite sets and relations. In other words, we interpret each combinator 
{{c :  b1 <-> b2}} as a \emph{relation} between the sets denoted
by {{b1}} and {{b2}} respectively. As we illustrate, this semantics allows
\emph{arbitrary} relations (including the empty relation and
relations that are not isomorphisms) to be represented as values in {{PiTF}}.
In Sec.~\ref{sec:cat} we discuss a more refined semantics that is
more suitable for richer categories and which only allows relations
representing \emph{isomorphisms} to be represented as values in~{{PiTF}}.

@@
